---
title: "Day 3 - Module 11 Gene Annotations"
author: "UM Bioinformatics Core"
date: "`r Sys.Date()`"
output:
        html_document:
            includes:
                in_header: header.html
            theme: paper
            toc: true
            toc_depth: 4
            toc_float: true
            number_sections: true
            fig_caption: true
            markdown: GFM
            code_download: true
---
<style type="text/css">
body{ /* Normal  */
      font-size: 14pt;
  }
pre {
  font-size: 12pt
}
code.r{
  font-size: 12pt;
}
</style>

<!--- Allow the page to be wider --->
<style>
    body .main-container {
        max-width: 1200px;
    }
</style>

<br>

# [Breakout Exercise 2] - Finding annotations

<br>


```{r Modules, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
library(DESeq2)
library(ggplot2)
library(tidyr)
library(dplyr)
library(matrixStats)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
load("rdata/RunningData.RData")
```


**15 Minutes**

<br>



<br>

### Instructions:

<br>

- One group member should share their screen in the breakout room. If nobody volunteers, a helper may randomly select someone.
- The group members should discuss the exercise and work together to find a solution.
- If there is time after a solution is found, allow all members to complete the exercise.

<br>

- Add the gene symbols stored in the `GeneKey` object to the DE results table `res_WT` and store the new table in an object named `res_WT_anno`.


> Hint:
> 
> Look at the documentation for either the `merge` function or the tidyverse `join` function for ideas on how to combine the `GeneKey` and `res_WT` tables
>

```{r CheckKey}
## check the biomaRt key we created to remember the structure
head(GeneKey)
```

<br>


<details>
    <summary>*Click for solution for adding annotations*</summary>
    First, create a new table called `res_WT_anno` that includes a column with the ENSEMBL ids named `genes` using the `mutate` function. Then use the `left_join` function to combine the `GeneKey` table with the `res_WT` DE results.
```{r updateDETableTidyr}
res_WT_anno <- as.data.frame(res_WT) %>%
  mutate(genes = row.names(res_WT)) %>%
  left_join(GeneKey, by =c("genes" = "ensembl_gene_id")) %>%
  relocate(c("genes", "external_gene_name")) # optionally, re-order columns to make output more readable

head(res_WT_anno)
```
    Alternatively, if you are more familiar with base functions:
```{r updateDETableBase}
res_WT_anno <- res_WT # copy table
res_WT_anno <- cbind(genes=row.names(res_WT_anno), res_WT_anno[ ,c(1:6)])
res_WT_anno <- as.data.frame(res_WT_anno)

# combine the two tables using the merge function (similar to join from `tidyverse`)
res_WT_anno <- merge(GeneKey, res_WT_anno, by.x = "ensembl_gene_id", by.y="genes", all.x = FALSE, all.y = TRUE)

head(res_WT_anno)
```
    Notice that not all genes were annotated with an ENSEMBl gene id or gene description. While we are able to annotate our results, this is a helpful reminder that gene symbol is often not a good unique identifier.

</details>

  
```{r WriteOut.RData, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
#Hidden code block to write out data for knitting
save.image(file = "rdata/RunningData.RData")
```


